\bf The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2022-02-03 paper/paper-jsdse/paper-jsdse.tex
pdf paper for jsdse

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this, is the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel. The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021).

This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University. The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average. An exploratory analysis reveals however that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.

This disparity between the average and the individual is a part of statistics that is given less attention. It is this disparity which makes this particular data set important as a textbook example. Textbook data sets have longevity if there is a unresolved mystery. The iris data (Andereson 1935) is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?

For the field of statistics, and data science by association, it is increasingly important to reach the individual. One might describe this as a divergence of purpose, statistics for public policy or statistics for the public. The two are not the same. As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual. For example, with the wages data, the message for the individual is that you are more than your demographic. The majority of people tracked are not like the average, the average is an anomaly. If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.

As textbook data though, the original data is outdated. The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published. Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students. Using tired old textbook data sets sets up a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) but wild data to textbook data.

@@

Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). (Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

>>>>

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, including hierarchical, mixed effects and multilevel. The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021).

The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average. An exploratory analysis reveals however that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages. The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.

This textbook data was used as an example in Ilk (2004), and was developed into a case study for use in teaching exploratory data analysis at CENSORED. The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average. An exploratory analysis reveals however that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.

This disparity between the average and the individual is a part of statistics, as a discipline, that is given less attention. It is this disparity which makes this particular data set important as a textbook example. Textbook data sets have longevity if there is a unresolved mystery. The iris data (Andereson 1935) is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

For the field of statistics, and data science by association, it is increasingly important to reach the individual. One might describe this as a divergence of purpose, statistics for public policy or statistics for the public. The two are not the same. As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual. For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic. The majority of people in the study do not have a pattern that is similar to the average. If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think. {[}ET: a related article: \url{https://www.smh.com.au/national/by-the-numbers--the-average-australian-doesnt-exist--not-a-single-one-of-us-is-normal-20150124-12xlh0.html}{]}

As a textbook data set though, it is outdated. The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published. Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students. Using tired old textbook data sets can imbue a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) wild data to textbook data.

@@

Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al. 2019). The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). (Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
